\bf Title here

### 2020-08-13 paper/paper.tex
describe some features of our optim prob; changes on the existing algorithms as per issue #5

\hypertarget{tour-optim}{%
\subsection{Optimisation problem and existing
algorithms}\label{tour-optim}}

@@

{[}you wanna give more information about the characteristics of the
optimisation problem before introducing existing algorithms {]}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

If the gradient information was available, the decreasing condition is

\hypertarget{difficulties}{%
\subsection{Difficulties}\label{difficulties}}

(feel this section should go after the general introduction of DFO and
PPGT)

Below are listed several issues in projection pursuit optimisation. Some
are general optimisation problems, while others are more specific too PP
optimisation.

\begin{itemize}
\item
  \emph{constraint optimisation}: our problem requires the project basis
  to be orthonormal
\item
  \emph{Multivariate optimisation}: In projection pursuit, the decision
  variable includes all the entries in the projection matrix, which is
  high-dimensional. Researchers would be better off if they could
  understand the relative position of different projection matrix in the
  high-dimensional space. (multivariate optimisation)
\item
  \emph{optimising non-smooth function}: When the objective function is
  non-differentiable, derivative information can not be obtained, which
  means traditional gradient- or Hessian- based methods are not
  feasible. Stochastic optimisation method could be an alternative to
  solve these problems. (derivative information is not available)
\item
  \emph{Finding global maximum}: Although finding local maximum is
  relatively easy with developed algorithms, it is generally hard to
  guarantee global maximum in a problem where the objective function is
  complex or the number of decision variables is large. Also, there are
  discussions on how to avoid getting trapped in a local optimum in the
  literature. (the exact \(\alpha\) that minimised
  \(f(x^{(k)}+ \alpha d^{(k)})\) can be hard to find or sometiems even
  impossible, imposing certain decrease, how every this doesn't
  guarantee global convergence)
\end{itemize}

Computational considerations:

\begin{itemize}
\item
  \emph{computation speed}: The optimisation procedure needs to be fast
  to compute since the tour produces real-time animation of the
  projected data.
\item
  \emph{consistency result in stochastic optimisation}: In stochastic
  algorithm, researchers usually set a seed to ensure the algorithm
  produces the same result for every run. This practice supports
  reproducibility, while less efforts has been made to guarantee
  different seeds will provide the same result.
\item
  \emph{role of interpolation in PP optimisation}: An optimisation
  procedure usually involves iteratively finding projection bases that
  maximise the index function, while tour requires geodesic
  interpolation between these bases to produce a continuous view for the
  users. It would be interesting to see if the interpolated bases could,
  in reverse, help the optimisation reach faster convergence.
\end{itemize}

\emph{Think about how does your package help people to understand
optimisation}

\begin{itemize}
\tightlist
\item
  diagnostic on stochastic optim
\item
  vis the progression of multi-parameter decision variable
\item
  understanding learning rate - neighbourhood parameter
\item
  understand where the local \& global maximum is found - trace plot -
  see if noisy function
\end{itemize}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

There are three existing algorithms.

\citet{posse1995projection} proposed a random search algorithm that
samples a candidate basis \(\mathbf{A}_{l}\) in the neighbourhood of the
current basis \(\mathbf{A}_{\text{cur}}\) by
\(\mathbf{A}_{l} = \mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{rand}\),
where \(\alpha\) controls the radius of the sampling neighbourhood and
\(\mathbf{A}_{rand}\) is a randomly generated matrix with the same
dimension as \(\mathbf{A}_{\text{cur}}\). The optimiser keeps sampling
bases near the current basis until it finds one with higher index value
than the current basis and then outputs it for guided tour to construct
the interpolation path. A new round of search continues to find a better
basis after the interpolation finishes. The halving parameter \(c\) with
default value of 30 is designed to adjust the searching neighbourhood
\(\alpha\). When the search needs to sample more than \(c\) number of
basis to find an outputted basis, the neighbourhood parameter \(\alpha\)
will be reduced by half in the next iteration. If the optimiser can't
find a better basis within the maximum number of tries \(l_{\max}\), the
algorithm stops. The algorithm is summarised in Algorithm
\ref{random-search} for one iteration.

\begin{algorithm}
\SetAlgoLined
  \SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}_{\text{cur}}$; The index function: $f$}
    \Output{The target basis: $\mathbf{A}_{l}$}
  initialisation\;
  Set $l = 1$ and $c = 0$\;
  \While{$l < l_{\max}$}{
    Generate $\mathbf{A}_{l} = \mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}$ ensuring $\mathbf{A}_{l}$ is orthonormal\;
    Compute $I_{l}  = f(\mathbf{A}_{l})$\;
    \eIf{$I_{l} > I_{\text{cur}}$}{
      \KwRet{$\mathbf{A}_{l}$} \;
      }{
      $c = c + 1$\;
      }
    $l = l + 1$\;
  }
  \caption{random search}
  \label{random-search}
\end{algorithm}

\citet{cook1995grand} explained the use of a gradient ascent
optimisation with an assumption that the index function is continuous
and differentiable. Since some indices could be non-differentiable, the
derivative is replaced by a pseudo-derivative of evaluating five
randomly generated directions in a tiny nearby neighbourhood. Once the
best direction is determined, the optimiser finds the best projection
\(\mathbf{A}_{**}\) on the geodesic between the current basis
\(\mathbf{A}_{\text{cur}}\) and the best directional basis
\(\mathbf{A}_{*}\). If the percentage change in the index value between
\(\mathbf{A}_{**}\) and \(\mathbf{A}_{\text{cur}}\) is greater than a
threshold value, \(\mathbf{A}_{**}\) is outputted for the current
iteration to construct the interpolation or the algorithm repeats the
above steps until \(l_{\max}\) is reached and the searching terminates.
Algorithm \ref{search-geodesic} summarise the steps in geodesic search.

\begin{algorithm}
\SetAlgoLined
\SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}_{\text{cur}}$; The index function: $f$}
    \Output{The target basis: $\mathbf{A}_{**}$}
  initialisation\;
  Set $l = 1$\;
  \While{$l < l_{\max}$}{
    Generate ten bases in five random directions: $\mathbf{A}_{l}: \mathbf{A}_{l+9}$ within a small neighbourhood $\text{dist}$\;
    Find the direction with the largest index value: $\mathbf{A}_{*}$ where $l < * < l+ 9$\;
    Construct the geodesic $\mathcal{G}$ from $\mathbf{A}_{\text{cur}}$ to $\mathbf{A}_{*}$\;
    Find $\mathbf{A}_{**}$ on the geodesic $\mathcal{G}$ that has the largest index value \;
    Compute $I_{**} = f(\mathbf{A}_{**})$, $p_{\text{diff}} = (I_{**} - I_{\text{cur}})/I_{**}$\;
      \If{$p_{\text{diff}} > 0.001$}{
         \KwRet{$\mathbf{A}_{**}$} \;
      }
    $l = l + 1$\;
  }
  \caption{search geodesic}
  \label{search-geodesic}
\end{algorithm}

Simulated annealing
\citep[\citet{kirkpatrick1983optimization}]{bertsimas1993simulated} is a
non-derivative procedure based on a non-increasing cooling scheme
\(T(l)\). Given an initial \(T_0\), the temperature at iteration \(l\)
is defined as \(T(l) = \frac{T_0}{\log(l + 1)}\). The simulated
annealing algorithm works as follows. Given a neighbourhood parameter
\(\alpha\) and a randomly generated orthonormal basis
\(A_{\text{rand}}\), a candidate basis is sampled by
\(\mathbf{A}_{t+1} = (1 - \alpha)\mathbf{A}_{t} + \alpha \mathbf{A}_{\text{rand}}\).
If the index value of the candidate basis is larger than the one of the
current basis, the candidate basis is outputted. While if it is smaller,
the candidate has another chance of being accepted with probability
\(P= \min\left\{\exp\left[\frac{I_{cur} - I_{l}}{T(l)}\right],1\right\}\)
where \(I_{\cdot}\) denotes the index value of a given basis. The
algorithm can be summarised as in Algorithm \ref{simulated_annealing}.

\begin{algorithm}
\SetAlgoLined
\SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}_{\text{cur}}$; The index function: $f$; Initial temperature: $T_0$}
    \Output{The global object; The target basis: $\mathbf{A}_{l}$}
  initialisation\;
  Set $l = 1$\;
  \While{$l < l_{\max}$}{
    Generate $\mathbf{A}_{l} = (1 - \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}$ ensuring $\mathbf{A}_{l}$ is orthonormal \;
    Compute $I_{l} = f(\mathbf{A}_{l})$ and $T(l) = \frac{T_0}{\log(l + 1)}$\;
      \eIf{$I_{l} > I_{\text{cur}}$}{
        \KwRet{$\mathbf{A}_{l}$} \;
      }{
        Compute $P= \min\left\{\exp\left[\frac{I_{\text{cur}} -I_{l}}{T(l)}\right],1\right\}$\;
        Draw $U$ from a uniform distribution: $U \sim \text{Unif(0, 1)}$\;
        \If{$P > U$}{
           \KwRet{$\mathbf{A}_{l}$} \;
        }
      }
    $l = l + 1$\;
  }
  \caption{simulated annealing}
  \label{simulated_annealing}
\end{algorithm}

Notice that the three algorithms above constitute one search. The
outputted basis is then sent to construct the interpolation and one such
search-and-interpolate step is one round.

>>>>

\hypertarget{tour-optim}{%
\subsection{Optimisation problem}\label{tour-optim}}

@@

There are several features of this optimisation that are worth noticing.
First of all, it is a multivariate constraint optimisation problem.
Since the decision variables are the entries of a projection basis, it
is required to be orthonormal. It is also likely that the objective
function is non-differentiable or the gradient information is simply not
available. In this case, we will need to either use some approximation
of the gradient or turn to derivative free methods. Given the goal of
projection pursuit as finding the basis with the largest index value,
the optimisation problem need to be able to find the global maximum.
Along the way, locla maximum may also be of our interest since they
could present unexpected interesting projections. There is also one
computational consideration: the optimisation procedure needs to be easy
to compute since the tour animation is played in real-time.

\hypertarget{existing-algorithms}{%
\subsection{Existing algorithms}\label{existing-algorithms}}

Below we introduce three possible algorithms: \texttt{search\_better}
and \texttt{search\_better\_random} are derivative free methods that
sample candidate basis in the neighbourhood whilst
\texttt{search\_geodesic} approximates the gradient information and acts
as a non-derivative version of the gradient acsent.

\citet{posse1995projection} proposed a random search algorithm that
samples a candidate basis \(\mathbf{A}_{l}\) in the neighbourhood of the
current basis \(\mathbf{A}_{\text{cur}}\) by
\(\mathbf{A}_{l} = \mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{rand}\),
where \(\alpha\) controls the radius of the sampling neighbourhood and
\(\mathbf{A}_{\text{rand}}\) is a randomly generated matrix with the
same dimension as \(\mathbf{A}_{\text{cur}}\). The optimiser keeps
sampling bases near the current basis until it finds one with higher
index value than the current basis and then outputs it for guided tour
to construct the interpolation path. A new round of search continues to
find a better basis after the interpolation finishes. The halving
parameter \(c\) with default value of 30 is designed to adjust the
searching neighbourhood \(\alpha\). When the search needs to sample more
than \(c\) number of basis to find an accpeted basis, the neighbourhood
parameter \(\alpha\) will be reduced by half in the next iteration. If
the optimiser can't find a better basis within the maximum number of
tries \(l_{\max}\), the algorithm stops. The algorithm is summarised in
Algorithm \ref{random-search} for one iteration. {[}mention
orthonormalise to ensure the constaint is fulfilled; don't use
derivative information but a random search{]}

\begin{algorithm}
\SetAlgoLined
  \SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{$\mathbf{A}_{\text{cur}}$, $f$, $\alpha$, $l_{\max}$} 
    \Output{$\mathbf{A}_{l}$}
  initialisation\;
  Set $l = 1$ and $c = 0$\;
  \While{$l < l_{\max}$}{
    Generate $\mathbf{A}_{l} = \mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}$ and orthonormalise $\mathbf{A}_{l}$\;
    Compute $I_{l}  = f(\mathbf{A}_{l})$\;
    \eIf{$I_{l} > I_{\text{cur}}$}{
      \KwRet{$\mathbf{A}_{l}$} \;
      }{
      $c = c + 1$\;
      }
    $l = l + 1$\;
  }
  We repeat this loop and if $c > 30$, half the $\alpha$. Reset $c$ to zero.
  \caption{random search}
  \label{random-search}
\end{algorithm}

Simulated annealing
\citep[\citet{kirkpatrick1983optimization}]{bertsimas1993simulated}
modifies \texttt{search\_better} based on a non-increasing cooling
scheme \(T(l)\). Given an initial \(T_0\), the temperature at iteration
\(l\) is defined as \(T(l) = \frac{T_0}{\log(l + 1)}\). When a candidate
basis fails to have an index value larger than the current basis,
simulated annealing gives it a second chance to be accepted with
probabiliby
\[P= \min\left\{\exp\left[-\frac{I_{cur} - I_{l}}{T(l)}\right],1\right\}\]
where \(I_{\cdot}\) denotes the index value of a given basis. This
implementation allows the algorithm to jump out of a local maximum and
enables a more holistic search of the whole parameter space. This
feature is particularly useful when the dimension of the projected space
is smaller than the number of informative variables in the dataset
(i.e.~a one dimenional projection of the dataset with two informative
variables). The algorithm can be written as replacing line 5-11 of
Algorithm \ref{random-search} with Algorithm \ref{simulated_annealing}.

\begin{algorithm}
\SetAlgoLined
    Compute $I_{l} = f(\mathbf{A}_{l})$ and $T(l) = \frac{T_0}{\log(l + 1)}$\;
      \eIf{$I_{l} > I_{\text{cur}}$}{
        \KwRet{$\mathbf{A}_{l}$} \;
      }{
        Compute $P= \min\left\{\exp\left[-\frac{I_{\text{cur}} -I_{l}}{T(l)}\right],1\right\}$\;
        Draw $U$ from a uniform distribution: $U \sim \text{Unif(0, 1)}$\;
        \If{$P > U$}{
           \KwRet{$\mathbf{A}_{l}$} \;
        }
      }
  \caption{simulated annealing}
  \label{simulated_annealing}
\end{algorithm}

\citet{cook1995grand} used a gradient ascent algorithm with
pseudo-derivative. Instead of computing the actual gradient of the index
function with respect to the projection basis matrix,
\texttt{search\_geodesic} samples \(2n\) bases that are randomly
generated on a uniform open ball with the radius controlled by the
\(\delta\) parameter and finds the most promising one to construct the
geodesic direction as an approximation to the graident. The \(\delta\)
parameter is usually set to be tiny to ensure the geogesic direction is
a good local approximation to the graident. Once the searching direction
is determined, the optimiser finds the best projection
\(\mathbf{A}_{**}\) on the geodesic as the candidate.
\(\mathbf{A}_{**}\) is outputted for the current iteration if the
percentage change in the index value between \(\mathbf{A}_{**}\) and
\(\mathbf{A}_{\text{cur}}\) is greater than a threshold value or the
algorithm repeats the above steps until \(l_{\max}\) is reached and the
searching terminates. Algorithm \ref{search-geodesic} summarise the
steps in geodesic search.

\begin{algorithm}
\SetAlgoLined
\SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{$\mathbf{A}_{\text{cur}}$, $f$, $l_{\max}$, $n$, $\delta$}
    \Output{$\mathbf{A}_{**}$}
  initialisation\;
  Set $l = 1$\;
  \While{$l < l_{\max}$}{
    Generate $2n$ bases in $n$ random directions: $\mathbf{A}_{l}: \mathbf{A}_{l+9}$ within a small neighbourhood $\delta$\;
    Find the direction with the largest index value: $\mathbf{A}_{*}$ where $l < * < l+ 9$\;
    Construct the geodesic $\mathcal{G}$ from $\mathbf{A}_{\text{cur}}$ to $\mathbf{A}_{*}$\;
    Find $\mathbf{A}_{**}$ on the geodesic $\mathcal{G}$ that has the largest index value \;
    Compute $I_{**} = f(\mathbf{A}_{**})$, $p_{\text{diff}} = (I_{**} - I_{\text{cur}})/I_{**}$\;
      \If{$p_{\text{diff}} > 0.001$}{
         \KwRet{$\mathbf{A}_{**}$} \;
      }
    $l = l + 1$\;
  }
  \caption{search geodesic}
  \label{search-geodesic}
\end{algorithm}
